Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (made easy) with ANTsR

### 2014-02-15 paper.tex
ENH: Fixed figures.
ENH:  Reconfiguration.

%\title{Multivariate Neuroimage Analysis with \textit{ANTsR}: Application to Supervised Brain Tumor Segmentation using Random Forests}
\title{Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (made easy) with \textit{ANTsR}}

@@

\begin{abstract} 
Automated brain tumor segmentation in MRI is complicated by the lack of prior knowledge 
concerning tumor location, spatial extent, shape, possible displacement of normal 
tissue, and intensity signature, all of which limit 
the applicability of well-established techniques that have demonstrated good performance 
in related tasks such as normal brain tissue segmentation.  Due to this 
limitation, we localize our prior knowledge strategy to each subject through the 
use of subject-specific multimodal asymmetrical features which demonstrate excellent
discriminative properties within the random forest machine learning framework.  
This approach is made possible with symmetric multimodal brain templates which
are optimal in terms of both shape and multivariate intensity.  Further performance improvements are demonstrated by a successive application of random forest models for increasing levels of feature refinement 
 to which we refer as {\it concatenated random forests}.  Although the derived workflow involves many disparate components, the newly developed \textit{ANTsR} open source software package significantly facilitates the proposed methodology.  \textit{ANTsR}, a comprehensive interface of the popular Advanced Normalization Tools (ANTs) repository with the \textit{R} statistical project,
integrates well-performing image analysis tools such as image registration, segmentation, bias correction, and template building with access to advanced statistical and machine learning techniques as well as state-of-the-art visualization possibilities.  
The reported algorithmic framework was the top-performing entry in the MICCAI 2013 Multimodal Brain Tumor Segmentation challenge and, to our knowledge, is the only entry for both challenge years (2012 \& 2013) which has been made publicly available.
\end{abstract}

\begin{keyword}
advanced normalization tools \sep BRATS \sep brain tumor segmentation \sep \textit{R} project 
%% keywords here, in the form: keyword \sep keyword
\end{keyword}

@@

\section{Introduction}
Given the complexity of tumor growth and appearance and the need
for precise volumetric measurements for tumor characterization, 
much research has been invested in computational approaches for 
automated segmentation of tumor regions in MR images.  Several 
approaches have been previously proposed in the literature as
detailed in recent reviews \citep{angelini2007,bauer2013}.  
Any type of commensurable algorithmic evaluation, however, is
extremely problematic due to widely varying performance assessments
described in the corresponding publications, lack of publicly available 
evaluation data, and privately kept algorithmic instantiations.  
In response to these issues, the Multimodal Brain Tumor Segmentation 
(BRATS) challenge was initiated in 2012 (and continued in 2013) under 
the auspices of the Medical Image Computing and Computer Assisted 
Intervention Society in association with their annual international 
conference \citep{menze2014}.

@@

In line with our philosophy regarding open science \citep{tustison2013,ince2012}, 
we made the code available as open source%
\footnote{
https://github.com/ntustison/BRATS2013
}
and, to our knowledge, we are the only competitors out of the 20 or so teams 
that participated in either years' challenges to do this.  All code is based
on the well-vetted and open source Insight Toolkit (ITK)%
\footnote{
http://www.itk.org/
} 
of the National Institutes of Health and is available through our \textit{ANTsR} 
package.

@@

%Although random forests have been proposed previously in the 
%literature for, among other things, supervised brain segmentation (e.g. 
%\cite{geremia2011}), 
%we demonstrate that \textit{ANTsR} significantly facilitates the construction 
%of a fully functional, parallelizable workflow for such a task and 
%that the performance of our proposed framework has proven extremely effective in 
%actual practice, specifically
% in the recent MICCAI 2013 Multimodal Brain Tumor Segmentation
%Challenge.%
%\footnote{
%http://martinos.org/qtim/miccai2013/
%}
%\begin{figure}
%  \centering
%  \includegraphics[width=90mm]{Figures/award.pdf}
%  \caption{First place certificate from the MICCAI 2013 BRATS challenge.}
%  \label{fig:award}
%\end{figure}

@@

In order to better characterize deviations from normal brain shape 
and appearance, several image features were derived using symmetric 
population-specific multivariate templates \citep{avants2008,avants2010}.  
For normal neuroanatomy, the use of spatial prior information 
coupled with image normalization capabilities has proven useful 
in producing improved segmentation results of ``expected'' brain tissue
such as cerebrospinal fluid, gray matter, and white matter 
\citep[e.g.,][]{ashburner1997}.  In contrast, accommodating spatial 
priors to model the presence of a possible tumor and its constituent tissue 
components is difficult. However, since the normal brain 
exhibits a bilaterally symmetric organization, we can 
use the presence of asymmetries to differentiate abnormal 
brain tissue (i.e. tumor).  A similar motivation prompted
the identification of the mid-sagittal plane of symmetry \citep{prima2002}
for feature generation in MS lesion \citep{geremia2011} and 
tumor \citep{geremia2012} identification.  However, this
approach does not take into account the displacement of 
normal tissue due to tumor growth causing the mid-sagittal plane
to deform from its planar structure (cf Figure \ref{fig:asymmetry} (a)).

To take into account these potential asymmetries, we 
require a data set with the same modalities as dictated
by the subject image acquisition protocol.  Although it
would be preferable to build population-specific multivariate
templates from normal data using the same acquisition 
parameters, we substituted publicly available data

A recent neuroimaging reproducibility study
by Landman et al. resulted in an open data cohort of 21
normal individuals, each imaged twice, comprising several
modalities including ASL, FLAIR, DTI, fMRI, T1, and T2 
\citep{landman2011}.  These data (known as the
``MMRR'' data set) were selected for deriving
a multivariate template due to its public availability and
inclusion of several modalities (even permitting future 
incorporation of modalities 
not currently included with the BRATS challenge into our 
segmentation framework).

As detailed in \cite{avants2008,avants2010}, 
given $K$ image modalities for $N$ subjects,  
${\mathbf I} = \{I_1,I_2,\ldots, I_K\}$, multivariate 
template construction iterates between optimizing the set 
of diffeomorphic transforms between the subjects and the 
template, 
$\left\{\left(\phi_1,\phi_1^{-1}\right),\ldots,\left(\phi_N,\phi_N^{-1}\right)\right\}$ 
and constructing the 
optimal multivariate template appearance 
$\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$ to minimize the
following cost function:
\begin{align}
  \sum_{n=1}^N 
        \Bigg[ D &\left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right) \\ \nonumber 
        +& \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k\left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\Bigg].
\end{align}
$D$ is the diffeomorphic shape distance,
\begin{align}
D\left( \phi( \mathbf{x},0),\phi( \mathbf{x},1)\right) = \int_0^1 \| \nu(\mathbf{x},t)\|_L dt
\end{align}
dependent on the choice of linear operator, $L$, and $\nu$
is the velocity field
\begin{align}
\nu\left( \phi(\mathbf{x},t) \right) = \frac{d\phi(\mathbf{x},t)}{dt},\,\,\, \phi(\mathbf{x},0) = \mathbf{x}.
\end{align}
Each pairwise registration employing the similarity metric $\Pi_k$ can 
be assigned a relative weighting, $\lambda_k$, to weight a particular
modality's influence in the construction process.  Once the multivariate
template has converged (typically four iterations are used), we symmetrize
the template by flipping each image contralaterally and then running the
multivariate template construction a second time using only the multivariate
template and its symmetrical version.

In terms of implementation, this template building algorithm is 
encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
available in the ANTs repository, which permits parallel processing on
an individual workstation or on a large computational cluster.  In
Figure \ref{fig:symmetrictemplates} we show mid-axial slices from
the MMRR multivariate symmetrical template consisting of FA%
\footnote{
Although DTI-based images were not included in the
challenge data, such images have shown discriminative potential 
\citep[e.g.,][]{price2003} warranting investigation in our future
work.  
}, FLAIR,  
T1, and T2 components.  These images have been made publicly available%
\footnote{
http://figshare.com/articles/Kirby\_multivariate\_template/852989.
All 7 components are stored under the more informal ``Kirby'' moniker.
}
along with other processed images such as corresponding
tissue priors and structural labels.

{\color{red}{Need to describe the actual feature images}}

\begin{figure}[htb]
  \centering
    \includegraphics[width=85mm]{Figures/templateKirby.pdf}
  \caption{Multivariate symmetric template created from the MMRR 
           data described in \cite{landman2011}.  Shown are the
           (a) FA, (b) FLAIR, (c) MPRAGE, 
           and (d) T2 template components.
          }
  \label{fig:symmetrictemplates}
\end{figure}

\begin{figure}[!htb]
  \centering
    \includegraphics[width=85mm]{Figures/asymmetryFeatureImages2.pdf}
  \caption{ $S \underset{b}{\leftrightsquigarrow} \leftrightarrow T$
          }
  \label{fig:asymmetryFeatures}
\end{figure}

\subsection{Multi-Modal Feature Image Generation}

%For both training and prediction protocols, feature images are generated
%for each subject using the selected input MRI.  
Key to any supervised regression or classification protocol are the 
selected features for training and subsequent prediction.  Based on previous
work and our own experience, the following feature images are generated
for use in both training and prediction:
\begin{itemize}
  \item Per modality (FLAIR, T1, T1C, T2)
    \begin{itemize}
      \item First-order neighborhood statistical images:
            mean, variance, skewness, and entropy. 
            Neighborhood radius $\in \{1,3\}$.
    \item GMM (stage 1) and MAP-MRF (stage 2) posteriors: CSF, gray matter, white 
          matter, necrosis, edema, non-enhancing tumor and enhancing tumor (or a
          subset for the simulated data).
    \item GMM (stage 1) and MAP-MRF (stage 2) connected component geometry 
          features:  distance to tumor core label, volume, volume to surface area ratio, eccentricity, and elongation
    \item Template-based:  symmetric template difference and 
          contralateral difference with Gaussian smoothing ($\sigma = 4$mm).
    \end{itemize}
  \item Miscellaneous: normalized Euclidean distance based on cerebral mask,
    log Jacobian image, and  (T1 - T1C) difference image.
\end{itemize}
The generation of all feature images is performed using the available script 
{\tt createFeatureImages.sh} which also preprocesses the input images performing N4
bias correction \citep{tustison2010} and intensity rescaling to $[0,1]$.  Note that 
it is easily expandable to include 
additional modalities and/or neighborhood and smoothing parameters.
In Figure \ref{fig:featureImages}
we provide sample mid-axial slices from the BRATS\_HG0301
data set from the Challenge cohort.

@@

\subsection{Supervised Brain Tumor Segmentation with Concatenated Random Forests}
\subsection{\textit{ANTsR}:  An ANTs/\textit{R} Interface}
\subsection{MICCAI BRATS 2013 Brain Tumor Data}

@@

\subsection{\textit{ANTsR}:  An ANTs/\textit{R} Interface}

@@

ANTs (Advanced Normalization Tools) was built, originally, to provide 
high performance image registration for medical image analysis
\citep{avants2008a} and based upon the mature Insight Toolkit (ITK)
sponsored by the National Institutes of Health.  Since then, ANTs has grown to include 
several robust medical image analysis solutions including bias 
correction \citep{tustison2010}, $n$-tissue multivariate segmentation 
\citep{avants2011}, template construction \citep{avants2010}, and cortical 
thickness estimation \citep{das2009} (many of which have been
introduced into ITK partially in an attempted leveraging of Linus's Law---``Given enough eyeballs, all bugs are shallow'').  
However, in the evolution of the toolkit, it became clear 
that robust statistical machinery was lacking for making inferences regarding
the data produced during the course of ANTs processing.  \textit{ANTsR} was developed
specifically to provide an interface between ANTs, a 
powerful neuroimaging toolkit for producing reliable imaging data 
transformations, and the \textit{R} project%
\footnote{
http://www.r-project.org
}
for statistical computing and visualization thus providing a complete
set of tools for multivariate neuroimage analysis \textit{ANTsR} intends to provide a modern framework for medical analytics, with a focus on imaging-assisted prediction and statistical power.

Careful consideration of available statistical software 
led to the adoption of \textit{R} to complement ANTs quantification resulting in the
\textit{ANTsR} package.
%The R project for statistical computing,%
%\footnote{
%www.r-project.org
%}
% or more compactly
%`R', is an environment for statistical computation
%and data visualization.  
\textit{R}'s open source code base, reliable software testing and distribution strategies,
and add-on packages coupled with its rapidly growing 
community of developers and users has caused wide-scale
adoption within both academia and industry.

@@

The \textit{ANTsR} package is publicly available on the github project hosting service.%
\footnote{
\href{https://github.com/stnava/ANTsR}{https://github.com/stnava/ANTsR}
}
Prior to installation of \textit{ANTsR}, several external R packages
need to be installed including: \verb#Rcpp#, \verb#signal#, \verb#timeSeries#, 
\verb#mFilter#, \verb#doParallel#, \verb#robust#, \verb#magic#, \verb#knitr#, \verb#pixmap#, 
\verb#rgl#, and \verb#misc3d#.  
See \href{http://stnava.github.io/software/2014/01/08/antsr/}{http://stnava.github.io/software/2014/01/08/antsr/} for current status. 
Additionally, in order
to perform the supervised brain segmentation as described 
in later sections, one needs to also install the packages
\verb#randomForest#, \verb#snowfall#, \verb#rlecuyer#,
and \verb#ggplot2#.%
\footnote{
Packages are easily installed using the {\tt install.packages()} \textit{R} mechanism.
}

In addition to \textit{R} and the add-on packages previously mentioned, CMake is also 
required.  CMake%
\footnote{
http://www.cmake.org/
}
is an open source tool for the management and building of 
large-scale software projects.  It is used
to coordinate the downloading of external packages,
such as the Insight Toolkit (ITK)%
\footnote{
http://www.itk.org/
}
and ANTs.  Further instructions for download and
installation can be found on the \textit{ANTsR} github website.  Feel
free to contact the authors if installation trouble occurs.  We note
that \textit{ANTsR} is currently only tested on UNIX-alikes such OSX and Ubuntu
operating systems.  \footnote{Windows installation should be possible
  but, to our knowledge, has not been attempted.}.

\subsubsection{Usage}
\textit{ANTsR} is intended to not only allow easy interchange between
medical imaging formats and \textit{R} but also to facilitate
reproducible scientific studies and the type compilable analysis
articles that are fundamental to journals such as
\textit{Biostatistics}.  Both \verb#knitr# and \verb#sweave#
facilitate integration of \textit{R}-code with the LaTeX document
preparation system.

An additional motivation for our development of \textit{ANTsR} (and
hopefully its acceptance by the community) 
stems from the ability to couple ANTs core 
functionality, including IO tools such as \verb#antsImageRead#, 
with the large number of \textit{R} statistical and
visualization packages.  Due to this combination, several
functions have been easily created for such neuroimaging-specific 
tasks as fMRI/ASL data manipulation and analysis,
voxel and ROI-based  analyses,
%(e.g. \verb#filterMRIforNetworkAnalysis#, \verb#aslPerfusion#)
and connectivity visualization. % (e.g. \verb#plotBasicNetwork#).
The user help menu and documentation for the library  and its
constituent functions are invoked in the similar manner as other
\textit{R} libraries.

>>>>

%\title{Multivariate Neuroimage Analysis with ANTsR: Application to Supervised Brain Tumor Segmentation using Random Forests}
\title{Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (made easy) with ANTsR}

@@

\begin{abstract} 
Automated brain tumor segmentation in MRI is complicated by the lack of prior knowledge 
concerning tumor location, spatial extent, shape, possible displacement of normal 
tissue, and intensity signature, all of which limit 
the applicability of well-established techniques that have demonstrated good performance 
in related tasks such as normal brain tissue segmentation.  Due to this 
limitation, we localize our prior knowledge strategy to each subject through the 
use of subject-specific multimodal asymmetrical features which demonstrate excellent
discriminative properties within the random forest machine learning framework.  
This approach is made possible with symmetric multimodal brain templates which
are optimal in terms of both shape and multivariate intensity.  Further performance improvements are demonstrated by a successive application of random forest models for increasing levels of feature refinement 
 to which we refer as {\it concatenated random forests}.  Although the derived workflow involves many disparate components, the newly developed ANTsR open source software package significantly facilitates the proposed methodology.  ANTsR, a comprehensive interface of the popular Advanced Normalization Tools (ANTs) repository with the R statistical project,
integrates well-performing image analysis tools such as image registration, segmentation, bias correction, and template building with access to advanced statistical and machine learning techniques as well as state-of-the-art visualization possibilities.  
The reported algorithmic framework was the top-performing entry in the MICCAI 2013 Multimodal Brain Tumor Segmentation challenge and, to our knowledge, is the only entry for both challenge years (2012 \& 2013) which has been made publicly available.
\end{abstract}

\begin{keyword}
advanced normalization tools \sep BRATS \sep brain tumor segmentation \sep R project for statistical computing
%% keywords here, in the form: keyword \sep keyword
\end{keyword}

@@

\section{Introduction}
Given the complexity of tumor growth and appearance and the need
for precise volumetric measurements for tumor characterization, 
much research has been invested in computational approaches for 
automated segmentation of tumor regions in MR images.  Several 
approaches have been previously proposed in the literature as
detailed in recent reviews \citep{angelini2007,bauer2013}.  
Commensurable algorithmic evaluation, however, is
extremely problematic due to widely varying performance assessments
described in the corresponding publications, lack of publicly available 
evaluation data, and privately kept algorithmic instantiations.  
In response to these issues, the Multimodal Brain Tumor Segmentation 
(BRATS) challenge was initiated in 2012 (and continued in 2013) under 
the auspices of the Medical Image Computing and Computer Assisted 
Intervention Society in association with their annual international 
conference \citep{menze2014}.

@@

In line with our philosophy regarding open science \citep{tustison2013,ince2012}, 
we made the code available as open source%
\footnote{
https://github.com/ntustison/BRATS2013
}
and, to our knowledge, we are the only competitors out of the 20 or so teams 
that participated in either years' challenges to do this.  All code is based
on the well-vetted and open source Insight Toolkit (ITK)%
\footnote{
http://www.itk.org/
} 
of the National Institutes of Health and is available through our ANTsR 
package.

@@

%Although random forests have been proposed previously in the 
%literature for, among other things, supervised brain segmentation (e.g. 
%\cite{geremia2011}), 
%we demonstrate that ANTsR significantly facilitates the construction 
%of a fully functional, parallelizable workflow for such a task and 
%that the performance of our proposed framework has proven extremely effective in 
%actual practice, specifically
% in the recent MICCAI 2013 Multimodal Brain Tumor Segmentation
%Challenge.%
%\footnote{
%http://martinos.org/qtim/miccai2013/
%}
%\begin{figure}
%  \centering
%  \includegraphics[width=90mm]{Figures/award.pdf}
%  \caption{First place certificate from the MICCAI 2013 BRATS challenge.}
%  \label{fig:award}
%\end{figure}

@@

In order to better characterize deviations from normal brain shape 
and appearance, several image features were derived using symmetric 
population-specific multivariate templates.  
For normal neuroanatomy, the use of spatial prior information 
coupled with image normalization capabilities has proven useful 
in producing improved segmentation results of ``expected'' brain tissue
such as cerebrospinal fluid, gray matter, and white matter 
\citep[e.g.,][]{ashburner1997}.  In contrast, accommodating spatial 
priors to model the presence of a possible tumor and its constituent tissue 
components is difficult. However, since the normal brain 
exhibits a bilaterally symmetric organization, we can 
use the presence of asymmetries to differentiate abnormal 
brain tissue (i.e. tumor).  A similar motivation prompted
the identification of the mid-sagittal plane of symmetry \citep{prima2002}
for feature generation in MS lesion \citep{geremia2011} and 
tumor \citep{geremia2012} identification.  However, this
approach does not take into account the displacement of 
normal tissue due to tumor growth causing the mid-sagittal plane
to deform from its planar structure (cf Figure \ref{fig:asymmetry}).

To take into account these potential asymmetries, we 
require a data set with the same modalities as dictated
by the subject image acquisition protocol.  Although it
would be preferable to build population-specific multivariate
templates from normal data using the same acquisition 
parameters \citep{avants2010}, we substituted well-known,
publicly available data since such normal data was not made available
for the BRATS 2013 challenge. 
A recent neuroimaging reproducibility study
by Landman et al. resulted in an open data cohort of 21
normal individuals, each imaged twice, comprising several
modalities including ASL, FLAIR, DTI, fMRI, T1, and T2 
\citep{landman2011}.  These data (known as the
``MMRR'' data set) were selected for deriving
a multivariate template due to its public availability and
inclusion of several modalities (even permitting future 
incorporation of modalities 
not currently included with the BRATS challenge into our 
segmentation framework).

\begin{figure*}[htb]
  \centering
    \includegraphics[width=160mm]{Figures/templateKirby.pdf}
  \caption{Multivariate symmetric template created from the MMRR 
           data \citep{landman2011}.  Of the seven modalities 
           comprising the set of study acquisitions, we illustrate the
           (a) FA, (b) FLAIR, (c) MPRAGE, and (d) T2 template components.
           The optimal transformation and averaging of the individual 
           subject images result in the asymmetric template represented at 
           the top of the middle column.  A horizontal reflection 
           perpendicular to the mid-sagittal
           plane, $T_R$, resulted in the contralateral counterpart represented
           at the bottom of the second column.  The final template seen on
           the right is a result of repeating the template construction using
           the two asymmetric templates as input.
          }
  \label{fig:symmetrictemplates}
\end{figure*}

As detailed in \cite{avants2008,avants2010}, 
given $K$ image modalities, ${\mathbf I} = \{I_1,I_2,\ldots, I_K\}$, 
for $N$ subjects,  multivariate 
template construction iterates between optimizing the set 
of diffeomorphic transforms between the subjects and the 
template, 
$\left\{\left(\phi_1,\phi_1^{-1}\right),\ldots,\left(\phi_N,\phi_N^{-1}\right)\right\}$ 
and constructing the 
optimal multivariate template appearance 
$\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$ to minimize the
following cost function:
\begin{align}
  \sum_{n=1}^N 
        \Bigg[ D &\left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right) \\ \nonumber 
        +& \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k\left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\Bigg].
\end{align}
$D$ is the diffeomorphic shape distance,
\begin{align}
D\left( \phi( \mathbf{x},0),\phi( \mathbf{x},1)\right) = \int_0^1 \| \nu(\mathbf{x},t)\|_L dt
\end{align}
dependent on the choice of linear operator, $L$, and $\nu$
is the velocity field
\begin{align}
\nu\left( \phi(\mathbf{x},t) \right) = \frac{d\phi(\mathbf{x},t)}{dt},\,\,\, \phi(\mathbf{x},0) = \mathbf{x}.
\end{align}
Each pairwise registration employing the similarity metric $\Pi_k$ can 
be assigned a relative weighting, $\lambda_k$, to weight a particular
modality's influence in the construction process.  Once the multivariate
template has converged (typically four iterations are used), we symmetrize
the template by flipping each image contralaterally and then running the
multivariate template construction a second time using only the multivariate
template and its symmetrical version.  This is illustrated conceptually in
Figure \ref{fig:symmetrictemplates}.

In terms of implementation, this template building algorithm is 
encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
available in the ANTs repository, which permits parallel processing on
an individual workstation or on a computational cluster.  In
Figure \ref{fig:symmetrictemplates} we show mid-axial slices from
the MMRR multivariate symmetrical template consisting of FA%
\footnote{
Although DTI-based images were not included in the
challenge data, such images have shown discriminative potential 
\citep[e.g.,][]{price2003} warranting investigation in our future
work.  
}, FLAIR,  
T1, and T2 components.  These images have been made publicly available
along with other processed images such as corresponding
tissue priors and structural labels.%
\footnote{
http://figshare.com/articles/Kirby\_multivariate\_template/852989.
All 7 components are stored under the more informal ``Kirby'' data set moniker.
}

After constructing the template offline, each data set is processed
by first registering the non-contrast T1-weighted image to the T1-weighted
component of the symmetric template.  To do this we use a recently developed 
SyN \citep{avants2011a} variant based on B-spline regularization 
which has demonstrated good performance in normal brain registration.
Good performance also extends to this pathological data scenario as 
indicated by visual inspection and the fact that the derived features
were amongst the most informative in our winning entry.

We denote the mapping from the subject, $S$, to the template, $T$,
space as $\phi_b: S \leftrightarrow \underset{b}{\leftrightsquigarrow}  T$
which consists of both affine and diffeomorphic components.  Transform
invertibility is essential for the features we describe below.  The first 
set of feature images is generated by warping the template components 
to the subject space and calculating the difference image (see the 
top portion of Figure \ref{fig:asymmetryFeatures}).  For example,
the T2 symmetric template difference image is calculated from
\begin{align}
  \mathrm{T2}\,\,\mathrm{symmetric}\,\,\mathrm{template}\,\,\mathrm{difference} = S_{\mathrm{T2}} - T_{\mathrm{T2}}\left(\phi_b^{-1}\right).
\end{align}
Note that the T1-weighted contrast and non-contrast images are paired with
the T1-weighted component of the symmetric template.  The second
set of template-based feature images is generated per modality 
as the difference image with the contralateral reflection.  This is achieved by calculating the reflection transform, $T_R$, in the symmetric template space and composing transforms
as follows to create the non-Euclidean contralateral modality counterpart:
\begin{align}
  S_{contralateral} = S\left( \phi_b^{-1}\left( T_R\left( \phi_b \right)\right)\right).
\end{align}
This process is illustrated in the bottom portion of Figure \ref{fig:asymmetryFeatures}. A final, related feature is the Jacobian
of the transform.  The basis for this final feature is that 
relatively larger Jacobian values might be indicative of tumor
expansion.

\begin{figure}[!htb]
  \centering
    \includegraphics[width=85mm]{Figures/asymmetryFeatureImages2.pdf}
  \caption{ Given the mapping between the template, $T$, and subject, $S$, domains ($\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$), various features can be calculated which demonstrate good discriminative qualities.  Feature images used are outlined in a dashed line.
Top:  Difference images with the symmetric multivariate template are created by warping the template to the subject space and performing a voxelwise subtraction from the original modality image.
Bottom:  Similarly, contralateral difference images are calculated from each modality per subject by generating the non-Euclidean contralateral image via the diffeomorphic transform $\phi_b$.  
          }
  \label{fig:asymmetryFeatures}
\end{figure}

\subsection{Other Multi-Modal Feature Image Generation}

In addition to the feature images outlined in the previous section, 
several others were generated as described as follows:
\begin{itemize}
  \item Per modality (FLAIR, T1, T1C, T2)
    \begin{itemize}
      \item First-order neighborhood statistical images:
            mean, variance, skewness, and entropy. 
            Neighborhood radius $\in \{1,3\}$.
    \item GMM (stage 1) and MAP-MRF (stage 2) posteriors: CSF, gray matter, white 
          matter, necrosis, edema, non-enhancing tumor and enhancing tumor (or a
          subset for the simulated data).
    \item GMM (stage 1) and MAP-MRF (stage 2) connected component geometry 
          features:  distance to tumor core label, volume, volume to surface area ratio, eccentricity, and elongation
    \item Template-based:  symmetric template difference and 
          contralateral difference with Gaussian smoothing ($\sigma = 4$mm).
    \end{itemize}
  \item Miscellaneous: normalized Euclidean distance based on cerebral mask,
    log Jacobian image, and  (T1 - T1C) difference image.
\end{itemize}
The generation of all feature images is performed using the available script 
{\tt createFeatureImages.sh} which also preprocesses the input images performing N4
bias correction \citep{tustison2010} and intensity rescaling to $[0,1]$.  Note that 
it is easily expandable to include 
additional modalities and/or neighborhood and smoothing parameters.
In Figure \ref{fig:featureImages}
we provide sample mid-axial slices from the {\tt BRATS\_HG0301}
data set from the Challenge cohort.

@@

\subsection{Supervised Brain Tumor Segmentation with Concatenated Random Forests}
\subsection{ANTsR:  An ANTs/R Interface}
\subsection{MICCAI BRATS 2013 Brain Tumor Data}

@@

\subsection{ANTsR:  An ANTs/R Interface}

@@

ANTs (Advanced Normalization Tools) was built, originally, to provide 
high performance image registration for medical image analysis
\citep{avants2008a} and based upon the mature Insight Toolkit (ITK)
sponsored by the National Institutes of Health.  Since then, ANTs has grown to include 
several robust medical image analysis solutions including bias 
correction \citep{tustison2010}, $n$-tissue multivariate segmentation 
\citep{avants2011}, template construction \citep{avants2010}, and cortical 
thickness estimation \citep{das2009} (many of which have been
introduced into ITK partially in an attempted leveraging of Linus's Law---``Given enough eyeballs, all bugs are shallow'').  
However, in the evolution of the toolkit, it became clear 
that robust statistical machinery was lacking for making inferences regarding
the data produced during the course of ANTs processing.  ANTsR was developed
specifically to provide an interface between ANTs, a 
powerful neuroimaging toolkit for producing reliable imaging data 
transformations, and the R project%
\footnote{
http://www.r-project.org
}
for statistical computing and visualization thus providing a complete
set of tools for multivariate neuroimage analysis ANTsR intends to provide a modern framework for medical analytics, with a focus on imaging-assisted prediction and statistical power.

Careful consideration of available statistical software 
led to the adoption of R to complement ANTs quantification resulting in the
ANTsR package.
%The R project for statistical computing,%
%\footnote{
%www.r-project.org
%}
% or more compactly
%`R', is an environment for statistical computation
%and data visualization.  
R's open source code base, reliable software testing and distribution strategies,
and add-on packages coupled with its rapidly growing 
community of developers and users has caused wide-scale
adoption within both academia and industry.

@@

The ANTsR package is publicly available on the github project hosting service.%
\footnote{
\href{https://github.com/stnava/ANTsR}{https://github.com/stnava/ANTsR}
}
Prior to installation of ANTsR, several external R packages
need to be installed including: \verb#Rcpp#, \verb#signal#, \verb#timeSeries#, 
\verb#mFilter#, \verb#doParallel#, \verb#robust#, \verb#magic#, \verb#knitr#, \verb#pixmap#, 
\verb#rgl#, and \verb#misc3d#.  
See \href{http://stnava.github.io/software/2014/01/08/antsr/}{http://stnava.github.io/software/2014/01/08/antsr/} for current status. 
Additionally, in order
to perform the supervised brain segmentation as described 
in later sections, one needs to also install the packages
\verb#randomForest#, \verb#snowfall#, \verb#rlecuyer#,
and \verb#ggplot2#.%
\footnote{
Packages are easily installed using the {\tt install.packages()} R mechanism.
}

In addition to R and the add-on packages previously mentioned, CMake is also 
required.  CMake%
\footnote{
http://www.cmake.org/
}
is an open source tool for the management and building of 
large-scale software projects.  It is used
to coordinate the downloading of external packages,
such as the Insight Toolkit (ITK)%
\footnote{
http://www.itk.org/
}
and ANTs.  Further instructions for download and
installation can be found on the ANTsR github website.  Feel
free to contact the authors if installation trouble occurs.  We note
that ANTsR is currently only tested on UNIX-alikes such OSX and Ubuntu
operating systems.  \footnote{Windows installation should be possible
  but, to our knowledge, has not been attempted.}.

\subsubsection{Usage}
ANTsR is intended to not only allow easy interchange between
medical imaging formats and R but also to facilitate
reproducible scientific studies and the type compilable analysis
articles that are fundamental to journals such as
\textit{Biostatistics}.  Both \verb#knitr# and \verb#sweave#
facilitate integration of R-code with the LaTeX document
preparation system.

An additional motivation for our development of ANTsR (and
hopefully its acceptance by the community) 
stems from the ability to couple ANTs core 
functionality, including IO tools such as \verb#antsImageRead#, 
with the large number of R statistical and
visualization packages.  Due to this combination, several
functions have been easily created for such neuroimaging-specific 
tasks as fMRI/ASL data manipulation and analysis,
voxel and ROI-based  analyses,
%(e.g. \verb#filterMRIforNetworkAnalysis#, \verb#aslPerfusion#)
and connectivity visualization. % (e.g. \verb#plotBasicNetwork#).
The user help menu and documentation for the library  and its
constituent functions are invoked in the similar manner as other
R libraries.

As mentioned earlier, we have made this entire framework
available as open source.  In addition to the ANTsR repository
already on github which houses both ANTs and ANTsR functionality, 
we created a special github repository specifically for this work
containing figures, references, and text.%
\footnote{
\href{https://github.com/ntustison/ANTsAndArboles}{https://github.com/ntustison/ANTsAndArboles}
}
Also, we posted all
scripts (R, shell, and perl) used to coordinate the ANTsR processing 
including:
\begin{itemize}
  \item {\tt applyModel.R}:  applies a random forest model to a new 
  feature data set from a testing subject resulting in a set of probability
  images (one for each label).
  \item {\tt applyTumorSegmentationModel.sh}:  generates the new feature image set 
  from the testing MRI (by calling {\tt createFeatureImages.sh}).
  organizes the file names in a csv file (via {\tt createCSVFileFromModel.R}),
  and applies the random forest model using {\tt applyModel.R}. 
  \item {\tt applyTumorSegmentationModelForCohort.pl}:  Coordinates tumor 
  segmentation on the computational cluster for a given cohort.
  \item {\tt createCSVFileFromModel.R}:  organizes the set of feature image
  file names in a csv file for input into {\tt applyModel.R}.
  \item {\tt createFeatureImages.sh};  creates the set of feature images given
  a set of co-registered input MRI from a single subject. 
  \item {\tt createModel.R}:  creates a random forest model given the input csv 
  file of the feature image file names for all training data and set of truth
  label maps.
  \item {\tt plotVariableImportance.R}:  produces a plot of the importance 
  ({\tt MeanDecreaseAccuracy}) of each feature variable used 
  in constructing the model.
\end{itemize}
We also include a fully functional 2-D example which performs
both testing and training on sample challenge data.
